Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-03-09 curvature.tex
moved the remainder of proofs to the main text
integrate remaining 'cuttables' into text

The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
Selected proofs of our theorems and lemmas can be found in the appendix.
Proofs omitted for space can be found in~\citep{curvature_arxiv}.

@@

\begin{lem}[{Song~\citep{Song2003-gf}}]
\cuttable{    \pushQED{\qed}}
	\label{lem:degree_extremes}
	For a tree $T$ with $n$ leaves:
	\begin{enumerate}
		\item $\degree{T} = 3n^2 - 13n + 14$, if $T$ is a ladder tree,
		\item $\degree{T} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, if $T$ is a balanced tree, and
		\item  $3n^2 - 13n + 14 \le \degree{T} \le 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, otherwise.
	\end{enumerate}
\cuttable{    \popQED}
\end{lem}

We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
\begin{restatable}{lem}{degreemaxdelta}
	\label{lem:degree_max_delta}
	Let $T$,$S$ be trees with $n \ge 3$ leaves, and assume w.l.o.g. that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge 3/4$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lem}{degreechange}
\label{lem:degree_change}
Let $T$ and $S$ be trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

\end{restatable}

@@

\begin{restatable}{cor}{pairedneighbors}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
$$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
\end{restatable}

We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
\begin{restatable}{lem}{degreemaxdeltaadjacent}
	\label{lem:degree_max_delta_adjacent}
	Let $T$, $S$ be trees with $n \ge 3$ leaves, s.t. $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
	Then:
	\begin{enumerate}
		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lem}{sharedneighbors}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{restatable}

@@

\begin{restatable}{thm}{flatcurvature}
\label{thm:flat_curvature}
Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
Let $\{T_n \mid n > k\}$ be a sequence of trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the rSPR graph.
\end{restatable}

@@

\begin{restatable}{lem}{curvaturedistancebound}
	\label{lem:curvature_distance_bound}
	Let $T$ and $S$ be two trees. Then:
	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
\end{restatable}

\begin{restatable}{lem}{maxadjacentcurvature}
	\label{lem:max_adjacent_curvature}
	The maximum curvature between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
\end{restatable}
This bound is tight and has been verified computationally for $n \le 7$.

@@

\begin{restatable}{lem}{curvaturelowerbound}
\label{lem:curvature_lower_bound}
	The curvature between adjacent trees with $n$ leaves is at least
	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
\end{restatable}

@@

\begin{restatable}{lem}{asymptotic}
\label{lem:asymptotic}
	Let $T$ and $S$ be trees with $n$ leaves.
	Then:
	\begin{enumerate}
		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lem}{mhcurvature}
	\label{lem:mh_curvature}
	Let $T$ and $S$ be trees with $n$ leaves. Then:
	\begin{align*}
	\curvature{T,S} - \frac{1}{3\dspr{T,S}}
	&\le \curvature{\MH; T,S} \\
	\curvature{\MH; T,S}
	&\le \curvature{T,S} + \frac{1}{3\dspr{T,S}},\text{ and }
\end{align*}
	$$\curvature{T,S} - 1/6
	\le \curvature{\MH; T,S}
	\le \curvature{T,S} + 1/6 .$$ 
\end{restatable}

@@

In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\citep{Whidden2015-yi}.
Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to significant ``bottlenecks'' like those we have observed in real data.
In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
%\cuttable{Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).}
% We will likely follow previous work \citep{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \citep{felsenstein1981evolutionary} requires both branch length and topology.

Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\citep{Song2006-xe}, graphs of BEAST~\citep{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\citep{Gusfield2002-il} that can be expressed as trees.
%{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
%We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
%We will also explore random walks on ranked trees~\citep{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \citep{Drummond2012-ek} that infer rooted ``time-trees.''
%Random walks on other discrete structures such as partitions~\citep{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}

\section{Acknowledgements}
The authors would like to thank Alex Gavruskin, Vladimir Minin, and Bianca Viray for helpful discussions.
They are also grateful to the authors of the SAGE and GAP4 software, especially Alexander Hulpke.

@@

\appendix

\section{Selected Proofs}

\cuttable{
\degreemaxdelta*
\begin{proof}
	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3, \\
	\end{align*}
	which is greater than 3/4 when $n \ge 3$.
    Similarly for (ii):
	\begin{align*}
		{\Delta}N &= \degree{S} - \degree{T} \\
		&\le\ (4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}) \\
		&\ \ \ \ \ \ \ \ - (3n^2 - 13n + 14) \\
		&\le\ (4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
		&=\ n^2 - 5n + 6.
	\end{align*}
\end{proof}
}

\cuttable{
\degreechange*
\begin{proof}
The set of permissible rSPR moves changes in four different ways due to the movement of $R$:
(i) subtrees that include nodes on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
(ii) subtrees that include nodes on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
(iii) $R$'s parent subtree may now be moved into $U$, and
(iv) $R$'s parent subtree may no longer be moved into $V$.
No additional moves are introduced or blocked by the original rSPR operation on $R$.

Thus the difference in rSPR degree is $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
\end{proof}
}

\cuttable{
\pairedneighbors*
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
\end{proof}
}

\degreemaxdeltaadjacent*
\begin{proof}
	We first prove (i).
	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) then maximally balancing the terms in the product $ka$ as follows.

	If the subtree of $k$ leaves is moved to the root, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf to balance the product $ka$ at the cost of increasing $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	Let ${\Delta}N = \degree{S} - \degree{T}$.
	\cuttable[
		In both cases we have
		$${\Delta}N \le 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},$$
	]{
	If we move the additional leaf, we have:
	\begin{align*}
		{\Delta}N &\le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) \\
		&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
	\end{align*}
like before.
Similarly, if we do not move the additional leaf, we also have:
\begin{align*}
	{\Delta}N &\le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) \\
&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
\end{align*}
}
proving (i).

\sharedneighbors*
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$ because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Then leaves $r' \in R$, $q' \in Q$, and $u' \in U$, for some subtree $U$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

\flatcurvature*
\begin{proof}
Because $d(T_n, S_n) = 1$, we will prove the theorem by showing that the mass transport term $W_{1,n}$ sits between two bounds, each of which has limit 1 as $n$ goes to infinity.

>>>>

The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.

@@

\begin{lem}[{Song~\citep{Song2003-gf}}]
	\label{lem:degree_extremes}
	For a tree $T$ with $n$ leaves:
	\begin{enumerate}
		\item $\degree{T} = 3n^2 - 13n + 14$, if $T$ is a ladder tree,
		\item $\degree{T} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, if $T$ is a balanced tree, and
		\item  $3n^2 - 13n + 14 \le \degree{T} \le 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, otherwise.
	\end{enumerate}
\end{lem}

We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
\begin{lem}
	\label{lem:degree_max_delta}
	Let $T$,$S$ be trees with $n \ge 3$ leaves, and assume w.l.o.g. that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge 3/4$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{lem}
\begin{proof}
	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3, \\
	\end{align*}
	which is greater than 3/4 when $n \ge 3$.
    Similarly for (ii):
	\begin{align*}
		{\Delta}N &= \degree{S} - \degree{T} \\
		&\le\ (4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}) \\
		&\ \ \ \ \ \ \ \ - (3n^2 - 13n + 14) \\
		&\le\ (4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
		&=\ n^2 - 5n + 6.
	\end{align*}
\end{proof}

@@

\begin{lem}
\label{lem:degree_change}
Let $T$ and $S$ be trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

\end{lem}
\begin{proof}
The set of permissible rSPR moves changes in four different ways due to the movement of $R$:
(i) subtrees that include nodes on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
(ii) subtrees that include nodes on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
(iii) $R$'s parent subtree may now be moved into $U$, and
(iv) $R$'s parent subtree may no longer be moved into $V$.
No additional moves are introduced or blocked by the original rSPR operation on $R$.

Thus the difference in rSPR degree is $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
\end{proof}

@@

\begin{cor}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
$$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
\end{cor}
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
\end{proof}

We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
\begin{lem}
	\label{lem:degree_max_delta_adjacent}
	Let $T$, $S$ be trees with $n \ge 3$ leaves, s.t. $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
	Then:
	\begin{enumerate}
		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
	\end{enumerate}
\end{lem}
\begin{proof}
	We first prove (i).
	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) then maximally balancing the terms in the product $ka$ as follows.

	If the subtree of $k$ leaves is moved to the root, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf to balance the product $ka$ at the cost of increasing $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	Let ${\Delta}N = \degree{S} - \degree{T}$.
	If we move the additional leaf, we have:
	\begin{align*}
		{\Delta}N &\le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) \\
		&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
	\end{align*}
like before.
Similarly, if we do not move the additional leaf, we also have:
\begin{align*}
	{\Delta}N &\le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) \\
&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
\end{align*}
proving (i).

@@

\begin{lem}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lem}
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$ because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Then leaves $r' \in R$, $q' \in Q$, and $u' \in U$, for some subtree $U$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

@@

\begin{thm}
\label{thm:flat_curvature}
Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
Let $\{T_n \mid n > k\}$ be a sequence of trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the rSPR graph.
\end{thm}
\begin{proof}
Because $d(T_n, S_n) = 1$, we will prove the theorem by showing that the mass transport term $W_{1,n}$ sits between two bounds, each of which has limit 1 as $n$ goes to infinity.

@@

\begin{lem}
	\label{lem:curvature_distance_bound}
	Let $T$ and $S$ be two trees. Then:
	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
\end{lem}
\begin{proof}
	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} - 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
	The lower bound follows similarly.
\end{proof}

\begin{lem}
	\label{lem:max_adjacent_curvature}
	The maximum curvature between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
\end{lem}
\begin{proof}
The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap and all other tree pairs are at distance 1.
By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
The amount of overlapping mass in the shared neighbors of $T$ and $S$ is thus $\frac{6n-17}{\max(\degree{T},\degree{S})}$.
The minimum mass transfer cost is thus $1 - \frac{6n-17}{\max(\degree{T},\degree{S})}$.
This is minimized when $\degree{T}=\degree{S}$ are as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.

The maximum curvature is thus $1 - \frac{\degree{T} - (6n-17)}{\degree{T}} = \frac{6n - 17}{\degree{T}}= \frac{6n-17}{3n^2-13n+14}$.
\end{proof}

This bound is tight and has been verified computationally for $n \le 7$.

@@

\begin{lem}
\label{lem:curvature_lower_bound}
	The curvature between adjacent trees with $n$ leaves is at least
	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
\end{lem}
\begin{proof}
    In light of Corollary \ref{cor:paired_neighbors}, the optimal mass transport cost is maximized (and therefore curvature minimized) across adjacent trees $T$ and $S$ by a combination of two effects: trees that cannot be paired at distance $1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
		As we will show, these effects can be optimized simultaneously.
    To bound these effects, let $m$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
    We can bound the mass transport cost from above by $1 + 2m$ because pairs of neighbors of adjacent trees are at most distance 3 apart.
	This gives a lower bound of $1 - (1 + 2m) / 1) = -2m$ on the curvature.

@@

\begin{lem}
\label{lem:asymptotic}
	Let $T$ and $S$ be trees with $n$ leaves.
	Then:
	\begin{enumerate}
		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
	\end{enumerate}
\end{lem}
\begin{proof}
	We first prove the lower bound in the uniform case, that is $\curvature{T,S} \le \ric{T,S}$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy uniform case with parameter $p$.
	Recall that $\curvature{T,S} = \curvature[1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p) \, \dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
	Then:
	\begin{align*}
		\frac{\curvature[p]{T,S}}{p} &= \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p \\
		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
		&= \frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p} \\
		&= 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&= \curvature{T,S}.
	\end{align*}

@@

\begin{lem}
	\label{lem:mh_curvature}
	Let $T$ and $S$ be trees with $n$ leaves. Then:
	\begin{align*}
	\curvature{T,S} - \frac{1}{3\dspr{T,S}}
	&\le \curvature{\MH; T,S} \\
	\curvature{\MH; T,S}
	&\le \curvature{T,S} + \frac{1}{3\dspr{T,S}},\text{ and }
\end{align*}
	$$\curvature{T,S} - 1/6
	\le \curvature{\MH; T,S}
	\le \curvature{T,S} + 1/6 .$$ 
\end{lem}
\begin{proof}
	We first prove the lower bound.
	By Lemma~\ref{lem:degree_max_delta_adjacent}, the quotient of degrees for two adjacent trees $\ge \frac{5}{6}$.
	Thus, the Hastings ratio is always $\ge \frac{5}{6}$.
	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$, and $W_1'(T,S)$ the cost for the MH random walk.
	Moreover, let $m_T(z)$ and $m_S(w)$ be the mass assigned for the uniform random walk and $m'_T(z)$ and $m'_S(w)$ be the mass assigned for the MH random walk, for each vertex $z \in N(T)$ and $w \in N(S)$.
	We construct an upper bound on $W_1'(T,S)$ by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, from $T$ to a neighbor of $S$, or from a neighbor of $T$ to $S$.
	That is, for each $W_1$ assignment $\xi(z,w)$, we send $\xi'(z,w) = \xi(z,w) \min\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right)$ of the mass from $z$ to $w$.
	The remaining $\xi(z,w) - \xi'(z,w)$ of the mass is moved from $T$ to $S$, $T$ to $w$, and $z$ to $S$ in the respective proportions
	$\xi(z,w) \max\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right) - \xi'(z,w)$,
	$\xi(z,w) \min\left(0, \frac{m'_T(z)}{m_T(z)} - \frac{m'_S(w)}{m_S(w)}\right)$, and,
	$\xi(z,w) \min\left(0, \frac{m'_S(w)}{m_S(w)} - \frac{m'_T(w)}{m_T(w)}\right)$.
	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
	Moreover, the affected mass must be moved through at most two additional trees.
	Then, $W_1' \le W_1 + \frac{2}{6}$.
	We now have:
	\begin{align*}
		\curvature{\MH; T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}.
	\end{align*}

	We obtain the upper bounds similarly to the lower bounds, by observing that the affected at most $\frac{1}{6}$ of the mass may move through at most two fewer trees (i.e. directly between $T$ and $S$ rather than a pair of neighbors at distance $\dspr{T,S} + 2$ from each other).
	Again, this is at most one fewer tree when $\dspr{T,S}=1$.
\end{proof}

@@

In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\citep{Whidden2015-yi}.
Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to significant ``bottlenecks'' like those we have observed in real data.
In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).
One avenue of exploration will be to follow previous work \citep{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \citep{felsenstein1981evolutionary} requires both branch length and topology.

One could extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, many graph structures remain to be explored.
In particular, the case of unrooted SPR, which is also commonly used in phylogenetic search, is an obvious next step.
One could also explore random walks on ranked trees~\citep{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \citep{Drummond2012-ek} that infer rooted ``time-trees.''
Random walks on other discrete structures such as partitions~\citep{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.

\section{Acknowledgements}
The authors would like to thank Alex Gavruskin, Vladimir Minin, and Bianca Viray for helpful discussions.
They are also grateful to the authors of the SAGE and GAP4 software, especially Alexander Hulpke.
A preliminary version of this work appeared in the 2016 Proceedings of the Thirteenth Workshop on Analytic Algorithmics and Combinatorics (ANALCO).
The authors would like to thank the conference organizers, reviewers, and attendees for helpful comments and discussions that greatly improved this full version of our work.

@@
